Comparing variations in sediment transport capacity between subglacial and subaerial systems

### 2023-03-10 theory.tex
intro update

  Glaciers expel large quantities of sediment from their termini \citep{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
  Recent papers examining increased sediment discharge form glacierized regions at both the glacier and regional scale \citep{costa2017,delaney2020,li2022,vergara2022}, create an imperative to examine the processes controlling sediment discharge from glacierized catchments. 
\ian{more details about paper... why have they not examined the processes more?}

  Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from below glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}. 
  On shorter time periods, pressurized water transports available sediment from the glacier \citep{walder1994,creyts2013,beaud2018b}, should enough sediment be present below the glacier (i.e. in a transport-limited regime).

  In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \citep{shields1936,meyer1948,engelund1967} and the width of the channel $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:tau_t}
      \tau_t \propto w_c\, v^2,
    \end{equation}
  \end{linenomath*}
  % 
  where $w_c$ is the width of the channel.
  Following the conservation of mass, the velocity of the water flowing through a rectangular channel is given as 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:v}
      v = \frac{Q_w}{S},
    \end{equation}
  \end{linenomath*}
  % 
  where $Q_w$ is water discharge,  and $S$ is the channel's cross-sectional area.

  In subaerial channels, with open channel flow, the cross-sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \citep{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross-sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width  and water velocity $v$.

@@

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.65\linewidth]{Cartoon.pdf}
      \caption{Cartoon of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$ , and subaerial ,$v_f$ , systems are shown by arrow length marks at the channel boundary. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$.
        Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
      \label{fig:cartoon}
    \end{figure}
  \end{center}

>>>>

  Glaciers expel large quantities of sediment from their termini \citep{hallet1996}.
  Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
  Increases in sediment transport  have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021} and the Andes \citep{vergara2022}.
  In some of these regions, increased water discharge and glacier melt have been attributed to the greater sediment transport \citep{bendixen2017,costa2017,li2021}.
  In other environments, enhanced access of melt water to subglacial sediment at high elevations was attributed as the cause of increased sediment transport \citep{delaney2020,vergara2022}.
  The observed changes to these systems create an imperative to examine the processes controlling sediment discharge from glacierized catchments. 
  Yet, the relationship between water discharge and sediment transport in subglacial environments remains minimally discussed and not contrasted with the subaerial river systems that these glaciers feed.

  Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from below glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}. 
  On shorter time periods, pressurized water transports sediment from the glacier \citep{walder1994,creyts2013,beaud2018}, should enough sediment be present below the glacier (i.e. in a transport-limited regime).

  In a transport-limited regime, sediment discharge responds to the sediment transport capacity or the amount of sediment that could be carried by the water. Sediment transport capacity depends on the shear stress between water and sediment it flows  over \citep{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment. The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through at channel so that 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:tau_t}
      \tau_t \propto w_c\, v^2,
    \end{equation}
  \end{linenomath*}
  % 
  where $w_c$ is the width of the channel.
  Following the conservation of mass, the velocity of the water flowing through a rectangular channel is given as 
  \begin{linenomath*}
    \begin{equation}
      \label{eq:v}
      v = \frac{Q_w}{S},
    \end{equation}
  \end{linenomath*}
  % 
  where $Q_w$ is water discharge,  and $S$ is the channel's cross-sectional area.

  In subaerial channels, with open channel flow, the cross-sectional area $S$ of the channel evolves with changing water discharge $Q_w$  by changing the width of the channel along with the depth \citep{leopold1953}. During a flood with greater water discharge $Q_w$, for instance, the cross-sectional area $S$ increases as the channel widens and deepens, along with water velocity $v$. As a result, the shear stress across the channel $\tau_t$ will increase in response to both the channel width $w_c$  and water velocity $v$.

@@

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.65\linewidth]{Cartoon.pdf}
      \caption{Cartoon of the situation with various parameters. Water velocity magnitude in the subglacial, $v_g$ , and subaerial ,$v_f$ , systems are shown by arrow length. $S_g$ and $S_f$ represent the cross-sectional area in subglacial and subaerial channels. The subglacial channel widths $w_{cg}$ remain steady, while the subaerial channel widths evolve $w_{cf}$, denoted by thick black line. Shear stress $\tau$ is responsible for the mobilization of sediment and increases with the number of makers at the channel bed. 
        Note that in the subaerial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
      \label{fig:cartoon}
    \end{figure}
  \end{center}
